---
title: "Traits exploratory analysis"
author: "Noelia Rodríguez Pérez"
output:
  word_document: default
  html_document:
    df_print: paged
  pdf_document: default
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Directories setting up

```{r}
workingDir <- getwd()
dataDir <- file.path(workingDir, "data")
resultsDir <- file.path(workingDir,"out") 
```

## Load packages

```{r libraries, message=FALSE}
library(tidyverse)
library(ggpubr)
library(FactoMineR)
library(factoextra)
library(corrplot)
library(GGally)
library(cowplot)
library(ape)
library(nlme)
library(gt)
library(ggtree)
library(phytools)
library(RRphylo)
library(picante)
library(ggnewscale)
```

## Load data

```{r input data, message=FALSE}
# Import mammals LQ phenotype of sequenced spp
toga_spp <- read_tsv(file.path(dataDir,"LQ_TOGA_species.tsv"), col_names = T)

# Import mammalian all traits (only phenotypic info no sequences info)
mammalian_traits <- read_tsv(file.path(dataDir,"Mammalian_LQ_values.txt"), col_names = T)

# Import families df
spp_family <- read_csv(file.path(dataDir,"spp_family.csv"), col_names = T)

# Load phylogenetic tree 
primates_tree <- read.tree(file.path(dataDir,"science.abn7829_data_s4.nex.tree"))

#anage_primates <- read_tsv(file.path(dataDir,"anage_build15.txt"), col_names = T, 
#                  col_types = list(
#                    "References" = col_character(),   # Needs to be specified or else its interpreted as <int>
#                    "Sample size" = col_factor(c("tiny", "small", "medium", "large", "huge"), ordered = T),
#                    "Data quality" = col_factor(c("low", "questionable", "acceptable", "high"), ordered = T))) %>% 
#  filter(Order == 'Primates',
#    !is.na(`Adult weight (g)`),
#    !is.na(`Maximum longevity (yrs)`))
```

Mammalian exploratory analysis

```{r formatting data, message=FALSE}
# Select only primates
primates_lq <- toga_spp %>% 
  filter(ORDER == "Primates") %>%
  mutate(label = str_replace(SPECIES, " ", "_"),
         label = str_replace_all(label, "\\s\\w+", "")) %>%
  # Remove duplicated spp
  group_by(SPECIES) %>%
  filter(if (any(!is.na(LQ))) {  # If any non-NA LQ value exists in the group
      !is.na(LQ)           # Keep only the non-NA LQ row
    } else {row_number() == 1    # If all are NA, just keep one of them
    }) %>%
  ungroup() 

# Select only primates
primates_traits <- mammalian_traits  %>%
  filter(Order == "Primates") %>% 
  select(-c("Order tree", "color", "Order", "LQ"))

# Merge both phenotypic dfs to fill missing LQ info (we have 60 spp but 47 with LQ info)
primates_traits <- left_join(primates_lq, primates_traits, by = c("SPECIES" = "Scientific name"))

# check that for all missing LQ in our original database there was no LQ calculated info in mammalian_traits, so we remove that column 
#naLQ_recovered_bymerge <- primates_traits2 %>%
#  filter(is.na(LQ.x)) %>%
#  select(SPECIES,LQ.x,LQ.y)

# Save info obtained
#write.csv(naLQ_recovered_bymerge, file.path(resultsDir,"naLQ_recovered_bymerge.csv"), row.names = FALSE)

# Export spp names with info
write.csv(primates_traits$SPECIES, file.path(resultsDir,"spp_names.csv"), row.names = FALSE)

# Merge both dfs to add family info
primates_traits <- left_join(primates_traits, spp_family[, -1], by = c("SPECIES" = "query"))

# Imputation: Fill missing LQ with their same genus LQ mean
imputed_LQ <- primates_traits %>%
  mutate(genus = str_replace_all(SPECIES, "\\s\\w+", ""),
         imputed = ifelse(is.na(LQ), "Yes", "No")) %>%
  group_by(genus) %>%
  # if there's more than one row in the genus group and there is missing LQ, then impute
  mutate(impLQ = ifelse(n() > 1 & is.na(LQ), mean(LQ, na.rm = T), LQ)) %>%
  ungroup() %>%
  select(TOGA_ID, ASSEMBLY,ORDER,Family,label,`Common name`,genus, everything()) %>%
  filter(!is.na(LQ))

# Reorder non imputed df columns
primates_traits <- primates_traits %>%
  select(TOGA_ID, ASSEMBLY,ORDER,Family,label,`Common name`, everything()) %>%
  filter(!is.na(LQ))

# Fill phenotypic data from AnAge
#anage_primates <- anage_primates %>%
#  select(-c("Kingdom", "Phylum", "Class", "Order","Female maturity (days)","Male maturity (days)","Gestation/Incubation (days)","Weaning (days)" , "Litter/Clutch size","Litters/Clutches per year","Inter-litter/Interbirth interval", "Birth weight (g)", "Weaning weight (g)","Growth rate (1/days)", "Source", "Specimen origin", "Sample size", "Data quality", "IMR (per yr)" , "MRDT (yrs)", "Metabolic rate (W)", "Temperature (K)", "References" )) %>%
#  mutate(scientificname = paste(Genus, Species))

# Merge both dfs to add extra pheno info
#primates_traits <- left_join(primates_traits, anage_primates, by = c("SPECIES" = "scientificname"))

# check that for all missing body weight in our original database there was no body weigth calculated info in mammalian_traits, so we remove that column 
#nabodyweigh_recovered_bymerge <- primates_traits2 %>%
#  filter(is.na(`Adult weight (g).x`)) %>%
#  select(SPECIES,`Adult weight (g).x`,`Adult weight (g).y`,`Maximum longevity (yrs).x`,`Maximum longevity (yrs).y`)
#We don't recover extra info so we discart using AnAge

# Save new df with family info
#write.csv(primates_traits, file.path(dataDir,"primates_traits_family.csv"), row.names = FALSE)

# Find the common names between tree labels and trait df
common_names <- intersect(primates_tree$tip.label, primates_traits$label)

# Print alphabetically order different spp, to check for misspell names
sort(setdiff(primates_tree$tip.label, primates_traits$label))
sort(setdiff(primates_traits$label, primates_tree$tip.label))

# Prune tree, keeping the species from our trait dataframe
# Drop the tips that are NOT in common_names
pruned_primates.tree <- drop.tip(primates_tree, setdiff(primates_tree$tip.label, common_names))

# Match the order of pruned_primates.tree$tip.label
primates_traits <- primates_traits %>%
  mutate(ordering = match(label, pruned_primates.tree$tip.label)) %>%
  arrange(ordering)  %>%
  filter(!is.na(ordering)) # We keep only spp with representation in the tree

#Create variable called LQoutlier, where  low outliers are below Q1 − 1.5*IQR and high outliers are above Q3 + 1.5*IQR
Q1 <- quantile(primates_traits$LQ, probs = 0.25, na.rm = T)
Q3 <- quantile(primates_traits$LQ, probs = 0.75, na.rm = T)
IQR <- IQR(primates_traits$LQ, na.rm = T)
global_median <- median(primates_traits$LQ, na.rm = T)

primates_traits <- primates_traits %>% 
  mutate(LQoutlier = case_when(
           LQ < Q1 - (1.5*IQR) ~ "Low_outlier",
           LQ > Q3 + (1.5*IQR) ~ "High_outlier",
           TRUE ~ "No"),
         GD = LQ - global_median) %>% 
  group_by(Family) %>%  # Group by Family
  # Calculate each family median and standard deviation
  mutate(MedianFamLQ = median(LQ, na.rm = T),
         # Calculate each spp LQ family median differences 
         FD = LQ - MedianFamLQ)
# Alejandro's column outlier was depicting if spp was on the top50, with LQoutlier only homo sapiens is a high_outlier and was on top50 on alejandro's column
```

Trait distribution

```{r}
# Histogram
hist(primates_traits$LQ, xlab = "LQ", main = "Histogram of primates LQ")
dev.copy(png, file.path(resultsDir,"descriptive/LQ_distribution.png"))
dev.off()  # Close the PNG device 
```


Depict trait values on phylogeny

```{r non-circular plots}
primates_lq_tree2 <- primates_traits$LQ
names(primates_lq_tree2) <- primates_traits$label

png(file.path(resultsDir,"descriptive/dotTree.png"), width = 1200, height = 900, res = 300)
dotTree(pruned_primates.tree, primates_lq_tree2, length = 10, ftype = "i", cex = 0.7, fsize = 0.6)
dev.off()


tree_df <- ggtree::fortify(pruned_primates.tree)
tree_df$LQ <- primates_lq_tree2[tree_df$label]

global_median <- median(primates_traits$LQ, na.rm = T)
#pruned_primates.tree$LQ <- primates_lq_tree2
dotColor <- function(x) {
  ifelse(x > global_median, "blue", "lightblue") 
}
p <- ggtree(tree_df) + 
    geom_tiplab(offset = 1) + hexpand(0.4) + # expanding the plotting area so that tree tip labels fit well
  geom_tippoint(aes(color = dotColor(LQ), size = LQ)) +
    scale_color_identity() +
    theme(legend.position = "right")
# save plot
ggsave(plot = p, filename = file.path(resultsDir,"descriptive/ggtreedots.tiff"), dpi = 300, units = "in",width = 6, height = 6)

#obj <- contMap(pruned_primates.tree, primates_lq_tree2, plot = FALSE)
#plot(obj,lwd = 7,xlim = c(-0.2,3.6))
```
convert the phylo object to a tidy data frame, a tbl_tree object
```{r}
pruned_primates.tree2 <- as_tibble(pruned_primates.tree)

# mapping external data to the tree structure
pruned_primates.tree2 <- full_join(pruned_primates.tree2, primates_traits, by = 'label') 

# the tbl_tree object is converted to a treedata object
pruned_primates.tree2 <- tidytree::as.treedata(pruned_primates.tree2)
```


```{r families as tips plot}
# Exclude outlier from trait representation
primates_traits2 <- primates_traits
primates_traits2$LQ <- ifelse(primates_traits2$LQoutlier == "High_outlier", NA, primates_traits2$LQ)

primates_lq_tree2 <- primates_traits2 %>% 
  column_to_rownames(var = "label") %>% 
  select(LQ)

circ <- ggtree(pruned_primates.tree2, layout = "circular") +
  # using a different shape to represent outliers
  geom_tippoint(aes(color = Family, shape = ifelse(LQoutlier == "High_outlier", 4, 16), size = 3)) + 
  scale_shape_identity() +
  guides(size = "none") #remove size legend
p1 <- gheatmap(circ, primates_lq_tree2, offset = .8, width = .2, colnames_angle = 95, colnames_offset_y = .05) + 
  scale_fill_viridis_c(option = "A", name = "LQ\nvalue", na.value = "grey50") # Color tip labels. Using grey color for NA values (high outliers)

p1
# save plot
ggsave(plot = p1, filename = file.path(resultsDir,"descriptive/ggtree_tips.tiff"), dpi = 300, units = "in",width = 6, height = 6)
```


```{r families as edges}
circ <- ggtree(pruned_primates.tree2, layout = "circular") +
  geom_tree(aes(color = Family), size = 1.5)   # Color branches/edges

p2 <- gheatmap(circ, primates_lq_tree2, offset = .8, width = .2, colnames_angle = 95, colnames_offset_y = .05) + 
  scale_fill_viridis_c(option = "A", name = "LQ\nvalue", na.value = "grey50")  

p2
# save plot
ggsave(plot = p2, filename = file.path(resultsDir,"descriptive/ggtree_lines.tiff"), dpi = 300, units = "in",width = 6, height = 6)
```

```{r Global median differences}
# Exclude outlier from trait representation
primates_traits2 <- primates_traits
primates_traits2$GD <- ifelse(primates_traits2$LQoutlier == "High_outlier", NA, primates_traits2$GD)

primates_lq_tree2 <- primates_traits2 %>% 
  column_to_rownames(var = "label") %>% 
  select(GD)

circ <- ggtree(pruned_primates.tree2, layout = "circular") +
  geom_tree(aes(color = Family), size = 1.5)   # Color branches/edges

p3 <- gheatmap(circ, primates_lq_tree2, offset = .8, width = .2, colnames_angle = 95, colnames_offset_y = .05) + 
  scale_fill_gradient2(name = "GD value", limits = c(-0.9,0.9), na.value = "grey50")

p3
# save plot
ggsave(plot = p3, filename = file.path(resultsDir,"descriptive/ggtree_global_median_diff.tiff"), dpi = 300, units = "in",width = 6, height = 6)
```

```{r Family median differences}
# Exclude outlier from trait representation
primates_traits2 <- primates_traits
primates_traits2$FD <- ifelse(primates_traits2$LQoutlier == "High_outlier", NA, primates_traits2$FD)

primates_lq_tree2 <- primates_traits2 %>% 
  column_to_rownames(var = "label") %>% 
  select(FD)

circ <- ggtree(pruned_primates.tree2, layout = "circular") +
  geom_tree(aes(color = Family), size = 1.5)   # Color branches/edges

p4 <- gheatmap(circ, primates_lq_tree2, offset = .8, width = .2, colnames_angle = 95, colnames_offset_y = .05) + 
  scale_fill_gradient2(name = "FD value", limits = c(-1.05,1),  na.value = "grey50")

p4
# save plot
ggsave(plot = p4, filename = file.path(resultsDir,"descriptive/ggtree_fam_median_diff.tiff"), dpi = 300, units = "in",width = 6, height = 6)
```


```{r both median differences}
# Exclude outlier from trait representation
primates_traits2 <- primates_traits
primates_traits2$GD <- ifelse(primates_traits2$LQoutlier == "High_outlier", NA, primates_traits2$GD)
primates_traits2$FD <- ifelse(primates_traits2$LQoutlier == "High_outlier", NA, primates_traits2$FD)

primates_lq_tree2 <- primates_traits2 %>% 
  column_to_rownames(var = "label") %>% 
  select(GD,FD)

circ <- ggtree(pruned_primates.tree2, layout = "circular") +
  geom_tree(aes(color = Family), size = 1.5)   # Color branches/edges

p5 <- gheatmap(circ, primates_lq_tree2, offset = .8, width = .2, colnames_angle = 95, colnames_offset_y = .05) + 
  scale_fill_gradient2(name = "Median difference", limits = c(-1.05,1),  na.value = "grey50")

p5
# save plot
ggsave(plot = p5, filename = file.path(resultsDir,"descriptive/ggtree_both_median_diff.tiff"), dpi = 300, units = "in",width = 6, height = 6)
```

```{r phylogeny with imputed df trait}
# Find the common names between tree labels and trait df
common_names <- intersect(primates_tree$tip.label, imputed_LQ$label)

# Print alphabetically order different spp, to check for misspell names
sort(setdiff(primates_tree$tip.label, imputed_LQ$label))
sort(setdiff(imputed_LQ$label, primates_tree$tip.label))
# Prune tree, keeping the species from our trait dataframe
# Drop the tips that are NOT in common_names
pruned_primates.impt <- drop.tip(primates_tree, setdiff(primates_tree$tip.label, common_names))

# Match the order of pruned_primates.impt$tip.label
imputed_LQ <- imputed_LQ %>%
  mutate(ordering = match(label, pruned_primates.impt$tip.label)) %>%
  arrange(ordering)  %>%
  filter(!is.na(ordering))

pruned_primates.impt2 <- as_tibble(pruned_primates.impt)

# mapping external data to the tree structure
pruned_primates.impt2 <- full_join(pruned_primates.impt2, imputed_LQ, by = 'label') 

# the tbl_tree object is converted to a treedata object
pruned_primates.impt2 <- tidytree::as.treedata(pruned_primates.impt2)
# Exclude outlier from trait representation
imputed_LQ2 <- imputed_LQ  %>%
  group_by(Family) %>%  # Group by Family
  # Calculate each family median and standard deviation
  mutate(MedianFamLQ = median(LQ, na.rm = T))

imputed_LQ2 <- imputed_LQ2  %>%
  mutate(GD = impLQ - global_median,
         FD = impLQ - MedianFamLQ,
         GD = ifelse(imputed == "Yes" | label == "Homo_sapiens", NA, GD),
         FD = ifelse(imputed == "Yes" | label == "Homo_sapiens", NA, FD))

imputed_traits2 <- imputed_LQ2 %>% 
  column_to_rownames(var = "label") %>% 
  select(GD,FD)

circ <- ggtree(pruned_primates.impt2, layout = "circular") +
  geom_tree(aes(color = Family), size = 1.5)   # Color branches/edges

p6 <- gheatmap(circ, imputed_traits2, offset = .8, width = .2, colnames_angle = 95, colnames_offset_y = .05) + 
  scale_fill_gradient2(name = "Median difference", limits = c(-1.05,1),  na.value = "grey50")

p6
# save plot
ggsave(plot = p6, filename = file.path(resultsDir,"descriptive/ggtreeimputed_global_median_diff.tiff"), dpi = 300, units = "in",width = 6, height = 6)
```

```{r}
ggtree(pruned_primates.tree) + geom_tiplab() + 
  hexpand(.2) + geom_text(aes(label = ifelse(node <= 38, NA, node)), hjust = -.3) #To get the internal node number
```

```{r}
ggtree(pruned_primates.tree) + 
  geom_text(aes(label=node, subset=!isTip), hjust=-0.25) + 
  theme_tree2(legend.position="none") +
  labs(title="Phylogram with Branch Lengths in Million Years")
```


```{r}
ape::plot.phylo(pruned_primates.impt2)
ape::axisPhylo()  
mtext("Million Years before present", side =1, line =3)
```

```{r}
#pruned_primates.tree %>% 
#  as_tibble
```


```{r}
internal_nodes <- pruned_primates.tree2 %>% 
  as_tibble %>%
  select(parent) %>%
  distinct()
```

```{r}
tree <- as.phylo(pruned_primates.tree2)
groups_list <- caper::clade.members.list(tree, tips = FALSE,tip.labels = TRUE, include.nodes = TRUE)

# Convert list to dataframe
groups_df <- map_df(names(groups_list), function(x) {
  tibble(group = x, N_spp = length(groups_list[[x]]$tips), associated_spp = paste(groups_list[[x]]$tips, collapse = ", "),Intermediate_subnodes = paste(groups_list[[x]]$nodes[-1], collapse = ", "))
})

groups_plus4 <- groups_df %>%
  filter(N_spp >= 4)

# Export to CSV
write_csv(groups_plus4, file.path(resultsDir,"descriptive/groups_plus4.csv"))

```



Esto no sirve por ahora
```{r groups by phylogeny cuts, eval=FALSE}
# Cut tree at sequential points
my_cut <- seq(5, 55, by = 5)
Species_included <- list()
for (i in seq_along(my_cut)) {
  set.seed(123)
  tree <- cutPhylo(pruned_primates.tree, age = my_cut[i], keep.lineage = F)
  Species_included[[i]] <- tree$tip.label
}
# Name each element of the list using my_cut
names(Species_included) <- my_cut

# Convert list to data frame
tree_groups <- stack(Species_included)

# Rename columns
colnames(tree_groups) <- c("Species", "Myears_Cut")
# Add LQ info
tree_groups <- left_join(tree_groups, primates_traits, by = c("Species" = "label"))

N_Top_LQ <- c()
N_Bottom_LQ <- c()
global_median <- median(primates_traits$LQ, na.rm = T)

for (i in my_cut) {
  tmp <- tree_groups %>% 
    filter(Myears_Cut == i) %>%
    mutate(MediancutLongevity = median(LQ),
           sdcutLongevity = sd(LQ),
           spp_class = case_when(
             LQ < (global_median) & LQ < (MediancutLongevity - sdcutLongevity) ~ "Short LQ",
             LQ > (global_median) & LQ > (MediancutLongevity + sdcutLongevity) ~ "Long LQ",
             TRUE ~ "Not extreme"))

  N_Top_LQ <- c(N_Top_LQ, sum(tmp$spp_class == "Long LQ"))
  N_Bottom_LQ <- c(N_Bottom_LQ, sum(tmp$spp_class == "Short LQ"))
}

# The vectors N_Species_Increased_MLS and N_Species_Decreased_MLS now have the counts for each threshold.
data <- data.frame(
  my_cut,
  N_Top_LQ,
  N_Bottom_LQ)

# Plotting
p1 <- ggplot(data, aes(x = my_cut, y = N_Top_LQ)) +
  geom_line(color = "red") +
  labs(x = "Mill year cut", y = "n top LQ species") +
  theme_classic() + 
  labs(subtitle = "Global median + 1 SD as selected criteria")

# Plotting
p2 <- ggplot(data, aes(x = my_cut, y = N_Bottom_LQ)) +
  geom_line(color = "blue") +
  labs(x = "Mill year cut", y = "n bottom LQ species") +
  theme_classic()

combined_plot <- plot_grid(p1, p2,  ncol = 2) 
combined_plot 
# save plot
ggsave(plot = combined_plot, filename = file.path(resultsDir,"descriptive/extreme_milyears.tiff"), dpi = 300, units = "in",width = 6, height = 6)
```

```{r different threshold per cut, eval=FALSE}
# Create vector of thresholds
sd_threshold <- seq(0.05, 2, by = 0.05)
N_Top_LQ <- c()
N_Bottom_LQ <- c()

for (i in my_cut) {
  for (j in sd_threshold) {
    tmp <- tree_groups %>% 
    filter(Myears_Cut == i) %>%
    mutate(MediancutLongevity = median(LQ),
           sdcutLongevity = j*sd(LQ),
           spp_class = case_when(
             LQ < (global_median) & LQ < (MediancutLongevity - sdcutLongevity) ~ "Short LQ",
             LQ > (global_median) & LQ > (MediancutLongevity + sdcutLongevity) ~ "Long LQ",
             TRUE ~ "Not extreme"))
    
    N_Top_LQ <- c(N_Top_LQ, sum(tmp$spp_class == "Long LQ"))
    N_Bottom_LQ <- c(N_Bottom_LQ, sum(tmp$spp_class == "Short LQ")) 
  }
}

# The vectors N_Top_LQ and N_Bottom_LQ now have the counts for each threshold.
data <- data.frame(
  # Repeat each element of my_cut by the length of threshold
  rep_my_cut <- rep(my_cut, each = length(sd_threshold)),
  sd_threshold,
  N_Top_LQ,
  N_Bottom_LQ)

for (i in my_cut) {
  tmp <- data %>% 
    filter(rep_my_cut == i)
  # Plotting
  p1 <- ggplot(tmp, aes(x = sd_threshold, y = N_Top_LQ)) +
    geom_line(color = "red") +
    labs(x = "SD threshold", y = "n top LQ species") +
    theme_classic() + 
    labs(subtitle = paste("At",i,"mill years"))
  
  # Plotting
  p2 <- ggplot(tmp, aes(x = sd_threshold, y = N_Bottom_LQ)) +
    geom_line(color = "blue") +
    labs(x = "SD threshold", y = "n bottom LQ species") +
    theme_classic()
  
  combined_plot <- plot_grid(p1, p2,  ncol = 2) 
  # save plots
  ggsave(plot = combined_plot, filename = file.path(resultsDir,paste0("descriptive/extreme_threshold_",i,"mily.tiff")), dpi = 300, units = "in",width = 6, height = 6)
}
```


```{r LQ family distribution}
# Calculate global median and standard deviation
global_median <- median(primates_traits$LQ, na.rm = T)
global_sd <- sd(primates_traits$LQ, na.rm = T) 

primates <- primates_traits %>% 
  group_by(Family) %>%  # Group by Family
  # Calculate each family median and standard deviation
  mutate(MedianFamLQ = median(LQ),
         sdFamLQ = sd(LQ),
         # Classify each species in below or above 1 SD 
         spp_class = case_when(
           LQ < (global_median) & LQ < (MedianFamLQ - sdFamLQ) ~ "Below 1 SD",
           LQ > (global_median) & LQ > (MedianFamLQ + sdFamLQ) ~ "Above 1 SD",
           TRUE ~ "Within 1 SD")) %>% 
  ggplot(aes(x = reorder(Family, ordering), y = LQ, color = Family, fill = Family)) +
  geom_jitter(aes(shape = spp_class), size = 3) +
  # Represent each species with different shapes according the species clas they belong respect to their MLS
  scale_shape_manual(values = c("Below 1 SD" = 22, "Within 1 SD" = 21,"Above 1 SD" = 24)) +
  # Draw family median
  stat_summary(fun.y = median, geom = "errorbar",
               aes(ymax = ..y.., ymin = ..y..), linewidth = 1) +
  # Dray global median
  geom_hline(yintercept = median(primates_traits$LQ, na.rm = T), linetype = "dashed", color = "grey", linewidth = 1) +
  coord_flip() +
  labs(x = 'Family',
    y = 'Longevity Quotient') +
  theme_classic() + 
  labs_pubr() +
  theme(
    legend.position = "none",
    text = element_text(size = 18))

primates
# save plot
ggsave(plot = primates, filename = file.path(resultsDir,"descriptive/extreme_byfamily.tiff"), dpi = 300, units = "in",width = 6, height = 6)
```

Load mammalian all traits

```{r}
# Check continuous variables distribution
p <- primates_traits %>% 
  select(`Adult weight (g)`, `Maximum longevity (yrs)`, LQ) %>%
  ggpairs()
# save plot
ggsave(plot = p, filename = file.path(resultsDir,"descriptive/primates_ggpairs.tiff"), dpi = 300, units = "in",width = 6, height = 6)

# Select 2 variables, remove missing variables and calculate spearman correlation from them
cor_numer <- primates_traits %>% 
  select(`Adult weight (g)`, `Maximum longevity (yrs)`, LQ) %>% # select_if(is.numeric) 
  drop_na() %>%
  cor(., method = "spearman") #,use = "pairwise.complete.obs" )

# Specify the tiff device to save plot
tiff(filename = file.path(resultsDir,"descriptive/primates_corplot.tiff"), width = 6, height = 6, units = "in", res = 300)
#  Correlation plot
corrplot.mixed(cor_numer, number.digits = 2, tl.col = "black") # Set the color of text labels for correlation numbers
# variables outside diagonal tl.pos = "lt" 

# Close the device
dev.off()
```

Make scatter plot of weight vs MLS in between all mammals in database

```{r mammals, eval=FALSE}
# Plot the same way as before
p.mammalia <- anage_mammals %>% 
  ggplot(aes(x = `Adult weight (g)`, y = `Maximum longevity (yrs)`, color = Order)) +
  geom_point(size = 1) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(
    method = 'lm', 
    aes(`Adult weight (g)`, `Maximum longevity (yrs)`), 
    inherit.aes = FALSE,
    col = "black") +
  labs( x = 'Log10(Adult Weight)',
    y = 'Log10(Maximum Longevity)',
    caption = "Human Ageing Genomic Resources: AnAge Build 15") +
  theme_classic() + 
  labs_pubr() +
  theme(
    legend.position = "bottom",
    text = element_text(size = 16),
    plot.caption = element_text(size = 10))

p.mammalia
```

Make scatter plot of weight vs MLS in between all primates in database 
I select all primates, to make plots that I can compare with the ones from Alejandro

```{r primates, eval=FALSE}
# Plot
primates <- primates_traits %>% 
  ggplot(aes(x = `Adult weight (g)`, y = `Maximum longevity (yrs)`, color = Family)) +
  geom_point(size = 1) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(
    method = 'lm', 
    aes(`Adult weight (g)`, `Maximum longevity (yrs)`), 
    inherit.aes = FALSE,
    col = "black") +
  labs(x = 'Log10(Adult Weight)',
    y = 'Log10(Maximum Longevity)',
    caption = "Human Ageing Genomic Resources: AnAge Build 15") +
  theme_classic() + 
  labs_pubr() +
  theme(
    legend.position = "right",
    text = element_text(size = 18),
    plot.caption = element_text(size = 10))

# save plot
ggsave(plot = primates, filename = file.path(resultsDir,"descriptive/primates_scatter.tiff"), dpi = 300, units = "in",width = 6, height = 6)
```


```{r table}
# Clean and Summarize the Data
summary_data <- primates_traits %>%
  group_by(Family) %>%
  # Summarize data
  summarize(
    n = n(),
    MLS_Mean = mean(`Maximum longevity (yrs)`, na.rm = TRUE),
    MLS_SD = sd(`Maximum longevity (yrs)`, na.rm = TRUE),
    min_MLS = min(`Maximum longevity (yrs)`, na.rm = TRUE),
    max_MLS = max(`Maximum longevity (yrs)`, na.rm = TRUE),
    LQ_Mean = mean(LQ, na.rm = TRUE),
    LQ_SD = sd(LQ, na.rm = TRUE),
    min_LQ = round(min(LQ, na.rm = TRUE),1),
    max_LQ = round(max(LQ, na.rm = TRUE),1))  %>% 
  # Compute the MLS Limits
  mutate(MLS_Limits = paste(min_MLS, "-", max_MLS),
         LQ_Limits = paste(min_LQ, "-",max_LQ))
summary_data

# Use the gt function to create the table
table_display <- summary_data %>%
  select(Family,n,MLS_Mean,MLS_SD, MLS_Limits, LQ_Mean, LQ_SD, LQ_Limits) %>%
  gt() %>%
  tab_header(title = "Table. Mean MLS and LQ Computed Using All Records from our Primate Database") %>%
  
  # Formatting columns for better display
  fmt_number(columns = c("MLS_Mean", "MLS_SD", "LQ_Mean", "LQ_SD"), decimals = 2) %>%
  
  # Adding and removing table lines
  tab_options(
    table.border.top.style = "none",  # Remove the top border
    table.border.bottom.style = "solid",
    table.border.bottom.width = 2,
    column_labels.border.top.style = "double",
    column_labels.border.bottom.style = "double")

# Display the table
table_display
#gtsave(table_display, filename = file.path(resultsDir,"descriptive/primates_table.png"))
```

PGLS
I add some code to test it when I have the data

```{r PGLS}
# There are 2 missing values 
primates_traits <- primates_traits %>%
  rename(MaxLongevity = `Maximum longevity (yrs)`, AdultWeight = `Adult weight (g)`) %>%
  filter(!is.na(MaxLongevity),!is.na(AdultWeight))

# Find the common names between tree labels and trait df
common_names <- intersect(primates_tree$tip.label, primates_traits$label)
# Prune tree, keeping the species from our trait dataframe
pruned_primates.tree <- drop.tip(primates_tree, setdiff(primates_tree$tip.label, common_names))

bm.mam <- corBrownian(phy = pruned_primates.tree)

# Fit the PGLS model (comprobar codigo en capitulo q nos pasaron)
pgls_model <- gls(MaxLongevity ~ AdultWeight, data = primates_traits, correlation = bm.mam)

# View the results
plot(pgls_model)
summary(pgls_model)

# Get residuals from the PGLS model
pgls_residuals <- residuals(pgls_model)
#no parece que los residuos coincidan con los que tenia alejandro, aunque los de él fueron calculados para mamímeros, no solo para primates
primates_traits$pgls_residuals <- pgls_residuals
```

Maximum lifespan distribution in every primate family:
I create plots as in Alejandro's paper for primates to compare with his results once I have the data to perform the PGLS an use the residuals of the MLS regressed over body mass
Ask what the global criteria should be, if the species have to be only above the global median to be considered extreme long lived plus each family criteria or they need to be above global median plus above 1SD of the global median

```{r residuals family distributions}
# Calculate global median and standard deviation
global_median <- median(primates_traits$pgls_residuals)
global_sd <- sd(primates_traits$pgls_residuals) 

primates <- primates_traits %>% 
  group_by(Family) %>%  # Group by Family
  # Calculate each family median and standar deviation
  mutate(MedianFamLongevity = median(pgls_residuals),
         sdFamLongevity = sd(pgls_residuals),
         # Classify each species in below or above 1 SD 
         spp_class = case_when(
           pgls_residuals < (global_median) & pgls_residuals < (MedianFamLongevity - sdFamLongevity) ~ "Below 1 SD",
           pgls_residuals > (global_median) & pgls_residuals > (MedianFamLongevity + sdFamLongevity) ~ "Above 1 SD",
           TRUE ~ "Within 1 SD")) %>% 
  ggplot(aes(x = reorder(Family, ordering), y = pgls_residuals, color = Family, fill = Family)) +
  geom_jitter(aes(shape = spp_class),  size = 3) +
  # Represent each species with different shapes according the species clas they belong respect to their MLS
  scale_shape_manual(values = c("Below 1 SD" = 22, "Within 1 SD" = 21,"Above 1 SD" = 24)) +
  # Draw family median
  stat_summary(fun.y = median, geom = "errorbar",
               aes(ymax = ..y.., ymin = ..y..), linewidth = 1) +
  # Dray global median
  geom_hline(yintercept = median(primates_traits$pgls_residuals), linetype = "dashed", color = "grey", linewidth = 1) +
  coord_flip() +
  labs(x = 'Family',
    y = 'Maximum Lifespan residuals') +
  theme_classic() + 
  labs_pubr() +
  theme(legend.position = "none",
        text = element_text(size = 16))

primates
# save plot
ggsave(plot = primates, filename = file.path(resultsDir,"descriptive/primates_famdistributionr.tiff"), dpi = 300, units = "in",width = 6, height = 6)
```

```{r alejandro resi fam distributions}
# Calculate global median and standard deviation
global_median <- median(primates_traits$residuals)
global_sd <- sd(primates_traits$residuals) 

primates <- primates_traits %>% 
  group_by(Family) %>%  # Group by Family
  # Calculate each family median and standar deviation
  mutate(MedianFamLongevity = median(residuals),
         sdFamLongevity = sd(residuals),
         # Classify each species in below or above 1 SD 
         spp_class = case_when(
           residuals < (global_median) & residuals < (MedianFamLongevity - sdFamLongevity) ~ "Below 1 SD",
           residuals > (global_median) & residuals > (MedianFamLongevity + sdFamLongevity) ~ "Above 1 SD",
           TRUE ~ "Within 1 SD")) %>% 
  ggplot(aes(x = reorder(Family, ordering), y = residuals, color = Family, fill = Family)) +
  geom_jitter(aes(shape = spp_class),  size = 3) +
  # Represent each species with different shapes according the species clas they belong respect to their MLS
  scale_shape_manual(values = c("Below 1 SD" = 22, "Within 1 SD" = 21,"Above 1 SD" = 24)) +
  # Draw family median
  stat_summary(fun.y = median, geom = "errorbar",
               aes(ymax = ..y.., ymin = ..y..), linewidth = 1) +
  # Dray global median
  geom_hline(yintercept = median(primates_traits$residuals), linetype = "dashed", color = "grey", linewidth = 1) +
  coord_flip() +
  labs(x = 'Family',
    y = 'Maximum Lifespan residuals') +
  theme_classic() + 
  labs_pubr() +
  theme(legend.position = "none",
        text = element_text(size = 16))

primates
# save plot
ggsave(plot = primates, filename = file.path(resultsDir,"descriptive/primates_Alejandrofamdistributionr.tiff"), dpi = 300, units = "in",width = 6, height = 6)
```

Correspondence Analysis. This plot is obtained with factorminer

```{r PCA}
# First we calculate Principal component analysis with all numeric variables from anAge
res.pca <- primates_traits %>%
  select_if(is.numeric) %>%
  select(-ordering,-residuals,-dif) %>%
  PCA(., graph = FALSE)
```

The most important (or contributing) variables can be highlighted on the correlation plot as follow:

```{r correspondence analysis}
fviz_pca_var(res.pca, col.var = "contrib", repel = TRUE,
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07")) + theme(panel.grid = element_blank())
```

```{r, eval=FALSE}
var_quanti_cor <- as.data.frame(res.pca$var$cor)
var_quanti_cor
# ver de tree coverage, si me interesa
#info_cluster_elbow <- fviz_nbclust(res.pca, hcut, method = "silhouette", print.summary = TRUE )
```


si se hiciera por un factor, como el que usaron ellos de agrupar cada trait en 5 categorias, podría adaptar este código:

```{r, eval=FALSE}
#grp <- as.factor(cluster)
# Color variables by groups
#fviz_pca_var(res.pca, col.var = grp, 
#             palette = c("#0073C2FF", "#EFC000FF", "#868686FF"), legend.title = "Cluster")
```




```{r treecluster}
# save tree as phylo
pruned_primates.impt2 <- as.phylo(pruned_primates.impt2)
# export tree
write.tree(pruned_primates.impt2, file = file.path(resultsDir,"descriptive/treecluster/pruned_primates.impt2.newick"))

# calculate distance from root, for each clade (tips+nodes)
# distance from root = cumulative branch length from root to the clade
# Compute distance of all tips and nodes to root
phylodistances <- castor::get_all_distances_to_root(pruned_primates.impt2)

# discart distances from tips & sort them
keep <- seq_along(pruned_primates.impt2$tip.label)
phylodistances <- sort(phylodistances[-keep])

# remove duplicates
phylodistances <- unique(phylodistances)
write(phylodistances, file = file.path(resultsDir, "descriptive/treecluster/phylodistances.txt"))
```



```{bash}
# Variables definition
out_dir="out/descriptive/treecluster/" 

# Cut the tree using TreeCluster 
for threshold in $(cat ${out_dir}phylodistances.txt) 
do
  # Cut the tree each node distance and collect the clusters in separated files 
  TreeCluster.py \
  -i ${out_dir}pruned_primates.impt2.newick \
  -t ${threshold} --method root_dist -o ${out_dir}cut_${threshold}.clusters
done
```

